---
title: "MP4"
subtitle: "STAT-340 Applied Regression Methods"
author: "Yerim Oh"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Details

### Due Date

This assignment is due at 11:59 PM on the 19th of November (no guaranteed pre-midterm feedback if submitted during the grace period.)

### Grading

The problems will be graded for correctness. An emphasis will be placed on full explanations of your thought process. You usually won't need to write more than a few sentences for any given problem, but you should write complete sentences! Understanding and explaining the reasons behind your decisions is more important than making the "correct" decision.

### Collaboration

You are allowed to work with others on this assignment, but you must complete and submit your own write up. You should not copy large blocks of code or written text from another student.

### Load Packages

The following R code loads packages needed in this assignment.

```{r, warning=FALSE, message=FALSE, echo = F}
library(readr)
library(readxl)
library(dplyr)
library(ggplot2)
library(gridExtra)
library(caret)
library(purrr)
library(rpart)
```

\newpage

## Problem 1: Adapted from ISLR Example 4.6.

Suppose we collect data for a group of students in a statistics class with variables

-   $X_1$ = hours studied
-   $X_2$ = undergrad GPA
-   $Y$ = receive an A in this class ("Yes" or "No")

We fit a logistic regression model and produce estimated coefficients, $\hat{\beta}_0 = -6$, $\hat{\beta}_1 = 0.05$, and $\hat{\beta}_2 = 1$.

### (a) What is the interpretation of the coefficient estimate $\hat{\beta}_1 = 0.05$, in terms of the odds of getting an A?

For an increase of 1 hour of studying, we expect the odds of the student to receive an A in this class to increase by a factor of $\exp(0.05) = 1.05127$.

### (b) Estimate the *probability* that a student who studies for 40 hours and has an undergrad GPA of 3.5 gets an A in the class.

$$
\hat{Pr}(Y=1|X_1 = 40, X_2 = 3.5) = \frac{e^{-6+0.05\cdot40+1\cdot3.5}}{1+e^{-6+0.05\cdot40+1\cdot3.5}} = 0.3775
$$

### (c) Estimate the *probability* that a student who studies for 41 hours and has an undergrad GPA of 3.5 gets an A in the class.

$$
\hat{Pr}(Y=1|X_1 = 41, X_2 = 3.5) = \frac{e^{-6+0.05\cdot41+1\cdot3.5}}{1+e^{-6+0.05\cdot41+1\cdot3.5}} = 0.3894
$$

### (d) By using your answer to part (b) and the definition of odds, estimate the *odds* that a student who studies for the class for 40 hours and has an undergrad GPA of 3.5 gets an A in the class. Do this again for the *odds* that a student who studies for the class for 41 hours and has an undergrad GPA of 3.5 gets an A in the class, using your answer to part (c). Verify that the interpretation you gave in part (a) holds in this example.

-   estimated *odds* that a student who studies for the class for 40 hours and has an undergrad GPA of 3.5 gets an A in the class:$$
    \begin{aligned}
    \frac{p(Y=1|X_1 = 40, X_2 = 3.5)}{1- p(Y=1|X_1 = 40, X_2 = 3.5)}
    = \frac{0.3775407}{1- 0.3775407} &= 0.6065307 \\
    e^{-6+0.05\cdot40+1\cdot3.5} &= 0.6065307
    \end{aligned}
    $$

-   estimated *odds* that a student who studies for the class for 41 hours and has an undergrad GPA of 3.5 gets an A in the class:$$
    \begin{aligned}
    \frac{p(Y=1|X_1 = 41, X_2 = 3.5)}{1- p(Y=1|X_1 = 41, X_2 = 3.5)}
    = \frac{0.3893608}{1- 0.3893608} &= 0.6376282 \\
    e^{-6+0.05\cdot41+1\cdot3.5} &= 0.6376282
    \end{aligned}
    $$

-   $\frac{0.6376282}{0.6065307} = 1.051271 = \exp(0.05)$

Therefore, the interpretation in part (a) holds, because in a multiple logistic regression, the ratio represents the change in odds of the response variable for a one-unit increase in a explanatory variable, while holding all other variables in the model constant.

### (e) Suppose a student has an undergrad GPA of 3.5. How many hours would they need to study for us to estimate that there is a probability of 0.5 that they will get an A in the class?

$$
\begin{aligned}
\hat{Pr}(Y=1|X_1 = x, X_2 = 3.5)
&= \frac{e^{-6+0.05\cdot x+1\cdot3.5}}{1+e^{-6+0.05\cdot x+1\cdot3.5}} = 0.5 \\
&= \frac{e^{-2.5+0.05\cdot x}}{1+e^{-2.5+0.05\cdot }} \\
= \frac{e^{-2.5}\cdot e^{0.05\cdot x}}{1+e^{-2.5}\cdot e^{0.05\cdot x}}
&= \frac{1}{2} \\
e^{-2.5}\cdot e^{0.05\cdot x} &= \frac{1}{2}(1+e^{-2.5}\cdot e^{0.05\cdot x}) \\
2\cdot(e^{-2.5}\cdot e^{0.05\cdot x}) &= 1+e^{-2.5}\cdot e^{0.05\cdot x} \\
2\cdot(e^{-2.5}\cdot e^{0.05\cdot x}) - (e^{-2.5}\cdot e^{0.05\cdot x}) &= 1 \\
e^{-2.5}\cdot e^{0.05\cdot x} &= 1 \\
e^{0.05\cdot x} &= \frac{1}{e^{-2.5}} = e^{2.5} \\
0.05\cdot x &= 2.5 \\
x &= 50
\end{aligned}
$$

A student with an undergrad GPA of 3.5 needs to study 50 hours for a probability of 0.5 that they will get an A in the class.

\vspace{12pt}

## Problem 2: Metabolic syndrome

[Metabolic syndrome](https://www.mayoclinic.org/diseases-conditions/metabolic-syndrome/symptoms-causes/syc-20351916) is a cluster of conditions that occur together, increasing your risk of heart disease, stroke and type 2 diabetes. These conditions include increased blood pressure, high blood sugar, excess body fat around the waist, and abnormal cholesterol or triglyceride levels. For the purpose of this assignment, we want to explore the association of hypertension status, weight and total cholesterol with the result of screening fasting blood sugar. We have the following variables:

-   `cat_fbs` : fasting blood sugar coded as `Normal`, `DM` (diabetes mellitus), and `IFG` (impaired fasting glucose).
-   `totchol`: Total Cholesterol (mmol/L).
-   `hptc`: Hypertension Status (`Yes` or `No`).
-   `weight`: Body weight measures in kilogram.

```{r, echo = FALSE}
ms <- read_xlsx("ms.xlsx")
ms <- ms %>%
  mutate(cat_fbs = factor(cat_fbs, levels = c("Normal","IFG", "DM")))
```

### (a) Fit a model to explain the fasting blood sugar (`cat_fbs`) with the other variables. Print out the model summary.

```{r, warning=FALSE, message=FALSE, echo = F}
multinom_fit <- train(
  cat_fbs ~ .,
  data = ms,
  trace = FALSE,
  method = "multinom",
  trControl = trainControl(method = "none")
)
summary(multinom_fit)
```

### (b) Interprete the coefficients $\hat{\beta}_{IFG, totchol}$ and $\hat{\beta}_{DM, totchol}$ in context.

$\hat{\beta}_{IFG, totchol} = 0.2394$: For an increase of 1 mmol/L total cholesterol, we expect the odds of impaired fasting glucose versus normal fasting blood sugar to increase by a factor of $\exp(0.2394) = 1.2705$, while the other variables are held constant.

$\hat{\beta}_{DM, totchol} = 0.2772$: For an increase of 1 mmol/L total cholesterol, we expect the odds of diabetes mellitus versus normal fasting blood sugar to increase by a factor of $\exp(0.2772) = 1.3194$, while the other variables are held constant.

### (c) Calculate the probability of having impaired fasting glucose (`IFG`) when totchol = 6, hptyes = yes (hypertension), and weight = 75kg.

$$
\begin{aligned}
& Pr(\text{IFG} | \text{totchol} = 6, \text{hptyes} = 1, \text{weight} = 75) \\
&= \frac{e^{-5.1127+0.2394\cdot6+0.8672\cdot1+0.02199\cdot75}}
{1 + e^{-5.1127+0.2394\cdot6+0.8672\cdot1+0.02199\cdot75} +
e^{-4.9079+0.2772\cdot6+0.9000\cdot1+0.02268\cdot75}}\\
&= \frac{0.3135332}{1+0.3135332+0.525345} = 0.1705024
\end{aligned}
$$

### (d) Calculate the probability of a normal fasting blood sugar (`Normal`) when totchol = 6, hptyes = yes (hypertension), and weight = 75kg.

$$
\begin{aligned}
& Pr(\text{Normal} | \text{totchol} = 6, \text{hptyes} = 1, \text{weight} = 75) \\
&= \frac{1}
{1 + e^{-5.1127+0.2394\cdot6+0.8672\cdot1+0.02199\cdot75} +
e^{-4.9079+0.2772\cdot6+0.9000\cdot1+0.02268\cdot75}}\\
&= \frac{1}{1+0.3135332+0.525345} = 0.5438098
\end{aligned}
$$

### (e) \underline{Extra credit}: By definition, under multinomial regression, we have that $$\operatorname{odds}(Y=k \mid X)=\frac{p(Y=k \mid X)}{p(Y=K \mid X)}.$$

### Show that:

$$e^{\beta_{kr}}=\frac{\operatorname{odds}\left(Y=k \mid X_r+1, \boldsymbol{X}_{-r}\right)}{\operatorname{odds}\left(Y=k \mid X_r, \boldsymbol{X}_{-r}\right)},$$

### where $\operatorname{odds}(Y=k \mid X)=\frac{p(Y=k \mid X)}{p(Y=K \mid X)}$ and $\boldsymbol{X}_{-r}$ denotes all predictors except $X_r$.

$$
\begin{aligned}
\operatorname{odds}(Y=k \mid X) &= \frac{p(Y=k \mid X)}{p(Y=K \mid X)} \\
\\
&= \frac{ \frac{ e^{\beta_{k0} + \beta_{k1}x_1 + \cdots + \beta_{kp}x_p} }
{ 1 + \sum_{j=1}^{K-1} e^{\beta_{j0} + \beta_{j1}x_1 + \cdots + \beta_{jp}x_p} } }
{\frac{ 1 }
{ 1 + \sum_{j=1}^{K-1} e^{\beta_{j0} + \beta_{j1}x_1 + \cdots + \beta_{jp}x_p} } }
= \frac{ e^{\beta_{k0} + \beta_{k1}x_1 + \cdots + \beta_{kp}x_p} } {1} \\
\\
&= e^{\beta_{k0} + \beta_{k1}x_1 + \cdots + \beta_{kp}x_p}
\\
\\
\operatorname{odds}(Y=k \mid X_r, X_{-r})
&= e^{\beta_{k0} + \beta_{k1}x_1 + \cdots + \beta_{kr}x_r + \cdots + \beta_{kp}x_p} \\
\\
\operatorname{odds}(Y=k \mid X_r+1, X_{-r})
&= e^{\beta_{k0} + \beta_{k1}x_1 + \cdots + \beta_{kr}(x_r+1) + \cdots + \beta_{kp}x_p} \\
&= e^{\beta_{k0} + \beta_{k1}x_1 + \cdots + \beta_{kr}x_r + \beta_{kr} + \cdots + \beta_{kp}x_p} \\
&= (e^{\beta_{k0} + \beta_{k1}x_1 + \cdots + \beta_{kr}x_r + \cdots + \beta_{kp}x_p}) \cdot e^{\beta_{kr}} \\
\\
\frac{ \operatorname{odds}(Y=k \mid X_r+1, X_{-r}) }{ \operatorname{odds}(Y=k \mid X_r, X_{-r}) }
&= \frac{ (e^{\beta_{k0} + \beta_{k1}x_1 + \cdots + \beta_{kr}x_r + \cdots + \beta_{kp}x_p}) \cdot e^{\beta_{kr}} }
{ e^{\beta_{k0} + \beta_{k1}x_1 + \cdots + \beta_{kr}x_r + \cdots + \beta_{kp}x_p} } \\
&= e^{\beta_{kr}}
\end{aligned}
$$

\vspace{12pt}

## Problem 3: Breast cancer diagnosis (if covered on Friday 11/15)

This problem is adapted from an example in the book "Extending the Linear Model with R" by Julian J. Faraway. The data set `wbca`, read in below, comes from a study of breast cancer in Wisconsin. There are 681 cases of potentially cancerous tumors of which 238 are actually malignant. Determining whether a tumor is really malignant is traditionally determined by an invasive surgical procedure. The purpose of this study was to determine whether a new procedure called fine needle aspiration, which draws only a small sample of tissue, could be effective in determining tumor status.

```{r}
wbca <- faraway::wbca %>%
  mutate(Class =   ifelse(Class == 1, "Benign", "Malignant"))
```

### (a) Fit a multiple logistic regression where `Class` is the response variable and "Adhes + BNucl + Chrom + Epith + Mitos + NNucl + Thick + UShap" are predictors. Print out the model summary.

### (b) Using your fitted model, construct a confusion matrix. Find the classification accuracy, classification error rate, false positive rate, and true positive rate.

### (c) Suppose we change the cut off to 0.1, so that we will classify a tumor as malignant if the estimated probability that it is malignant is at least 0.1. Explain why someone might realistically use a cut off of 0.1 in the context of medical screening using this procedure.
